---
title: "Predictive mapping of wholesale crop prices for rural areas in Tanzania"
#author: "Madaga Lavinia, Jordan Chamberlin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
css: styles.css
---

<div class="custom-citation">
Suggested citation: 
Madaga Lavinia, Jordan Chamberlin, Bisrat Gebrekidan, Maxwell Mkondiwa, Nicholaus Kuboja, and Silva Joao Vasco, 2026. “Predictive mapping of wholesale crop prices for rural areas in Tanzania.” https://github.com/Analytics-for-Rural-Transformation-ART/Tanzania-Agricultural-Price-Predictions
</div>

# Introduction

In this notebook, we demostrate the process of estimating the prices of agricultural food commodities across space and time, on the basis of prices as observed at some market locations. Here we explore methods to model these prices, using monthly data from across Tanzania over the period May 2021-January 2026.

This document describes the process of fitting and cross validating random forest models to predict these prices. We use CPI, seasonality, accessibility and  bioclimatic covariates to fit our random forest model. For a given crop j, market i and month t, our model can be written as: 

$$P_{tij} = f(\text{Crop, Seasonality, CPI, Market Access, Climate, Pop Density, Rainfall Lag, } t) + \epsilon$$

We use four different cross validation srategies to separately evaluate pooled and crop specific data models.  The performance of the random forest models are evaluated using Root Mean Square Errors (RMSE) and R-Square as primary test statistics.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
```

# Load Libraries
```{r, message=FALSE}
library(geodata)
library(lubridate)
library(terra)
library(data.table)
library(randomForest)
library(httr)
library(caret)
library(Metrics)
library(pdp)
library(gridExtra)
library(stats)
library(dplyr)
library(stringr)
library(fields)
library(corrplot)
library(ggplot2)
library(tidyr)
library(gdistance)
library(sp)
library(raster)
library(foreign)
library(gdistance)
library(foreach)
library(doParallel)
library(tidyverse)
library(segmented)
library(modelbased)
library(parameters)
library(RColorBrewer)
library(effsize)
library(viridis)
library(scales)
library(patchwork)
```

# Data
This dataset contains price information for various crops across different regions and markets in Tanzania from May 2021 to January 2026. The data was acquired from Tanzania's Ministry of Industry and Trade - https://www.viwanda.go.tz/documents/product-prices-domestic.
```{r}
prices <- fread("Tanzania_Price_Data_AllCrops_with_Coordinates_5.csv")
dim(prices)
head(prices)
table(prices$Market)
sapply(prices, class)
```


```{r}
# Convert to date format
prices$Date <- lubridate::mdy(prices$Date)
```

## Basic Data preperation

Check the Region and Market names and coodinates.
```{r}
unique(prices[, c("Region", "Market", "Latitude", "Longitude")])
```

```{r}
setnames(prices, old = "Maize..min.price.", new = "mai.price.min")
setnames(prices, old = "Rice..min.price.", new = "ric.price.min")
setnames(prices, old = "Sorghum..min.price.", new = "sor.price.min")
setnames(prices, old = "Bulrush.Millet..min.price.", new = "bul.price.min")
setnames(prices, old = "Finger.Millet..min.price.", new = "fin.price.min")
setnames(prices, old = "Wheat..min.price.", new = "whe.price.min")
setnames(prices, old = "Beans..min.price.", new = "bea.price.min")
setnames(prices, old = "Irish.Potatoes..min.price.", new = "pot.price.min")
setnames(prices, old = "Maize..max.price.", new = "mai.price.max")
setnames(prices, old = "Rice..max.price.", new = "ric.price.max")
setnames(prices, old = "Sorghum..max.price.", new = "sor.price.max")
setnames(prices, old = "Bulrush.Millet..max.price.", new = "bul.price.max")
setnames(prices, old = "Finger.Millet..max.price.", new = "fin.price.max")
setnames(prices, old = "Wheat..max.price.", new = "whe.price.max")
setnames(prices, old = "Beans..max.price.", new = "bea.price.max")
setnames(prices, old = "Irish.Potatoes..max.price.", new = "pot.price.max")
```


```{r}
#convert prices to numeric 
prices$mai.price.min <- as.numeric(prices$mai.price.min)
prices$ric.price.min <- as.numeric(prices$ric.price.min)
prices$sor.price.min <- as.numeric(prices$sor.price.min)
prices$bul.price.min <- as.numeric(prices$bul.price.min)
prices$fin.price.min <- as.numeric(prices$fin.price.min)
prices$whe.price.min <- as.numeric(prices$whe.price.min)
prices$bea.price.min <- as.numeric(prices$bea.price.min)
prices$pot.price.min <- as.numeric(prices$pot.price.min)

prices$mai.price.max <- as.numeric(prices$mai.price.max)
prices$ric.price.max <- as.numeric(prices$ric.price.max)
prices$sor.price.max <- as.numeric(prices$sor.price.max)
prices$bul.price.max <- as.numeric(prices$bul.price.max)
prices$fin.price.max <- as.numeric(prices$fin.price.max)
prices$whe.price.max <- as.numeric(prices$whe.price.max)
prices$bea.price.max <- as.numeric(prices$bea.price.max)
prices$pot.price.max <- as.numeric(prices$pot.price.max)

sapply(prices, class)
```

```{r}
# convert to price per kg
prices$mai.price.min <- prices$mai.price.min/100
prices$ric.price.min <- prices$ric.price.min/100
prices$sor.price.min <- prices$sor.price.min/100
prices$bul.price.min <- prices$bul.price.min/100
prices$fin.price.min <- prices$fin.price.min/100
prices$whe.price.min <- prices$whe.price.min/100
prices$bea.price.min <- prices$bea.price.min/100
prices$pot.price.min <- prices$pot.price.min/100

prices$mai.price.max <- prices$mai.price.max/100
prices$ric.price.max <- prices$ric.price.max/100
prices$sor.price.max <- prices$sor.price.max/100
prices$bul.price.max <- prices$bul.price.max/100
prices$fin.price.max <- prices$fin.price.max/100
prices$whe.price.max <- prices$whe.price.max/100
prices$bea.price.max <- prices$bea.price.max/100
prices$pot.price.max <- prices$pot.price.max/100
```

```{r}
# calculate average of min and max
prices$mai.price <- (prices$mai.price.min + prices$mai.price.max) / 2
prices$ric.price <- (prices$ric.price.min + prices$ric.price.max) / 2
prices$sor.price <- (prices$sor.price.min + prices$sor.price.max) / 2
prices$bul.price <- (prices$bul.price.min + prices$bul.price.max) / 2
prices$fin.price <- (prices$fin.price.min + prices$fin.price.max) / 2
prices$whe.price <- (prices$whe.price.min + prices$whe.price.max) / 2
prices$bea.price <- (prices$bea.price.min + prices$bea.price.max) / 2
prices$pot.price <- (prices$pot.price.min + prices$pot.price.max) / 2
```

```{r}
#We can add dates by using the year and the month names
prices$Day   <- day(prices$Date)
prices$Month <- month(prices$Date)
prices$Year  <- year(prices$Date)
```

```{r}
# drop unneccessary columns
prices <- prices[,!c("mai.price.min", "mai.price.max",
                     "ric.price.min", "ric.price.max",
                     "sor.price.min", "sor.price.max",
                     "bul.price.min", "bul.price.max",
                     "fin.price.min", "fin.price.max",
                     "whe.price.min", "whe.price.max", 
                     "bea.price.min", "bea.price.max", 
                     "pot.price.min", "pot.price.max")]
```

```{r}
# We use the last data points of the month instead of using monthly averages

prices.monthly <- prices %>%
  arrange(Region, Market, Year, Month, Date) %>%  
  group_by(Region, Market, Year, Month) %>%
  summarise(
    Latitude  = last(Latitude),
    Longitude = last(Longitude),
    Reference_Date = last(Date),
    
    # Use the safe function for all crops
    mai.price = last(mai.price),
    ric.price = last(ric.price),
    sor.price = last(sor.price),
    bul.price = last(bul.price),
    fin.price = last(fin.price),
    whe.price = last(whe.price),
    bea.price = last(bea.price),
    pot.price = last(pot.price),
    .groups = "drop")

```

```{r}
# Make sure all price columns are numeric
price_cols <- c("mai.price","ric.price","sor.price","bul.price","fin.price","whe.price","bea.price","pot.price")
setDT(prices.monthly)
prices.monthly[, (price_cols) := lapply(.SD, as.numeric), .SDcols = price_cols]

# Then melt
prices.monthly.long <- melt(prices.monthly, 
                            id.vars=c('Region', 'Market', 'Month', 'Year', 'Latitude', 'Longitude'), measure.vars = price_cols)
```

```{r}
# rename columns
setnames(prices.monthly.long, old="variable", new="Crop")
setnames(prices.monthly.long, old="value", new="pkg")
```

```{r}
# replace crop names
prices.monthly.long[Crop == "mai.price", Crop := "Maize"]
prices.monthly.long[Crop == "ric.price", Crop := "Rice"]
prices.monthly.long[Crop == "sor.price", Crop := "Sorghum"]
prices.monthly.long[Crop == "bul.price", Crop := "B.Millet"]
prices.monthly.long[Crop == "fin.price", Crop := "F.Millet"]
prices.monthly.long[Crop == "whe.price", Crop := "Wheat"]
prices.monthly.long[Crop == "bea.price", Crop := "Beans"]
prices.monthly.long[Crop == "pot.price", Crop := "Potato"]
```

```{r}
# Reset the factor levels to updated levels
prices.monthly.long[, Crop := factor(Crop)]
# Check the unique values again
unique(prices.monthly.long$Crop)
```

```{r}
# generate dummies to use in place of factors (for later spatial predictions, which are struggling with factors)
prices.monthly.long[, maize   := ifelse(Crop == "Maize",1,0)]
prices.monthly.long[, rice    := ifelse(Crop == "Rice",1,0)]
prices.monthly.long[, sorghum := ifelse(Crop == "Sorghum",1,0)]
prices.monthly.long[, bmillet := ifelse(Crop == "B.Millet",1,0)]
prices.monthly.long[, fmillet := ifelse(Crop == "F.Millet",1,0)]
prices.monthly.long[, wheat   := ifelse(Crop == "Wheat",1,0)]
prices.monthly.long[, beans   := ifelse(Crop == "Beans",1,0)]
prices.monthly.long[, potato  := ifelse(Crop == "Potato",1,0)]
```

```{r}
prices.monthly.long[, jan := ifelse(Month == 1  , 1, 0)]
prices.monthly.long[, feb := ifelse(Month == 2  , 1, 0)]
prices.monthly.long[, mar := ifelse(Month == 3  , 1, 0)]
prices.monthly.long[, apr := ifelse(Month == 4  , 1, 0)]
prices.monthly.long[, may := ifelse(Month == 5  , 1, 0)]
prices.monthly.long[, jun := ifelse(Month == 6  , 1, 0)]
prices.monthly.long[, jul := ifelse(Month == 7  , 1, 0)]
prices.monthly.long[, aug := ifelse(Month == 8  , 1, 0)]
prices.monthly.long[, sep := ifelse(Month == 9  , 1, 0)]
prices.monthly.long[, oct := ifelse(Month == 10 , 1, 0)]
prices.monthly.long[, nov := ifelse(Month == 11 , 1, 0)]
prices.monthly.long[, dec := ifelse(Month == 12 , 1, 0)]
```

```{r}
# replace NaN with NAs in the price observations
prices.monthly.long[is.nan(pkg), pkg := NA]
# Remove observations with missing observations
prices.monthly.long <- na.omit(prices.monthly.long)
# head(prices.monthly.long, 5)
```

## Outlier Screening
```{r}
# Remove non-positive values (data errors)
prices.monthly.long <- prices.monthly.long[pkg > 0]

# Apply Statistical filter to remove extreme outliers by Crop
# We define an outlier as any value 3x IQR above Q3 or below Q1
prices.monthly.long[, c("Q1", "Q3") := .(quantile(pkg, 0.25, na.rm = TRUE), 
                                         quantile(pkg, 0.75, na.rm = TRUE)), by = Crop]

prices.monthly.long[, IQR_val := Q3 - Q1]
prices.monthly.long[, `:=`(lower_bound = Q1 - 3 * IQR_val, 
                           upper_bound = Q3 + 3 * IQR_val)]

# Identify rows to keep
prices.monthly.long <- prices.monthly.long[pkg >= lower_bound & pkg <= upper_bound]

# remove helper columns
prices.monthly.long[, c("Q1", "Q3", "IQR_val", "lower_bound", "upper_bound") := NULL]

# Check final counts per crop to ensure you still have ~10,137 observations
print(prices.monthly.long[, .N, by = Crop])
```

## Deflating Nominal Prices 
We use the Consumer Price Index (CPI) for Tanzania to convert nominal prices to real prices. We set May 2021 as our base period ($CPI_{base} = 104.22$). 
$$P_{real} = P_{nominal} \times \left( \frac{CPI_{base}}{CPI_{current}} \right)$$
```{r}
# Load  CPI data
cpi_tz <- read.csv("CPI- Tanzania.csv") 

# Join CPI to long format data
prices.monthly.long <- prices.monthly.long %>%
  left_join(cpi_tz, by = c("Year", "Month"))

# Calculate Real Price to a base of  May 2021 = 104.22
set(prices.monthly.long, j = "pkg_real", value = prices.monthly.long$pkg * (104.22 / prices.monthly.long$CPI))

# Add a check to ensure no NAs were introduced by the join
if(any(is.na(prices.monthly.long$pkg))) {
  warning("Some prices could not be deflated. Check if CPI data covers all Years/Months.")
}
```


```{r}
# bring in raster stack as predictors
geodata_path("C:/H/Tanzania Price data/Datasets/geodata")
# list.files("C:/H/Tanzania Price data/Datasets/geodata", recursive=TRUE)
```

```{r}
# tza0 <- gadm(country="TZA", level=0)
# tza1 <- gadm(country="TZA", level=1)
# tza2 <- gadm(country="TZA", level=2)
# tza3 <- gadm(country="TZA", level=3)

tza0 <- readRDS("C:/H/Tanzania Price data/Datasets/geodata/TRUE/gadm/gadm41_TZA_0_pk.rds")
tza1 <- readRDS("C:/H/Tanzania Price data/Datasets/geodata/TRUE/gadm/gadm41_TZA_1_pk.rds")
tza2 <- readRDS("C:/H/Tanzania Price data/Datasets/geodata/TRUE/gadm/gadm41_TZA_2_pk.rds")
tza3 <- readRDS("C:/H/Tanzania Price data/Datasets/geodata/TRUE/gadm/gadm41_TZA_3_pk.rds")

```

```{r}
# convert prices observations to vector for mapping
mypts <- vect(prices.monthly.long, geom=c("Longitude", "Latitude"), crs=crs(tza0), keepgeom=TRUE)

# see if these show up correctly
plot(tza1)
plot(mypts, col="Red", add=TRUE)
# text(mypts, label="Market")
```

```{r}
# Extract Region information from GADM
region_info <- terra::extract(tza1, mypts)
# head(region_info)

# merge district names to the dataset
prices.monthly.long$Region_GADM <- region_info$NAME_1
prices.monthly.long$District_GADM <- region_info$NAME_2

# head(prices.monthly.long)
# unique(prices.monthly.long[, c("Region", "Market", "Region_GADM")])

# Check if Region matches Region_GADM
prices.monthly.long$region_match <- prices.monthly.long$Region == prices.monthly.long$Region_GADM
# table(prices.monthly.long$region_match)

# Show mismatched rows
mis <- subset(prices.monthly.long, region_match == FALSE)
# unique(mis[, c("Region", "Market", "Region_GADM")])

# Songwe is mismatched, rename correctly
prices.monthly.long[Region_GADM == "Mbeya" & Region == "Songwe", Region_GADM := "Songwe"]
# unique(prices.monthly.long[, c("Region", "Market", "Region_GADM")])

# Check if Region matches Region_GADM
prices.monthly.long$region_match <- prices.monthly.long$Region == prices.monthly.long$Region_GADM
# table(prices.monthly.long$region_match)

# convert prices observations to vector for mapping
mypts <- vect(prices.monthly.long, geom=c("Longitude", "Latitude"), crs=crs(tza0), keepgeom=TRUE)
mkt_pts <- unique(mypts[, c("Market", "Latitude", "Longitude")])
# head(mkt_pts)
```


```{r}
# create reference raster
tza_extent <- ext(tza1) |> floor()
r <- crop(rast(res=1/12), tza_extent)
```

### Interpolate
```{r}
## Interpolate

#xy <- as.matrix(mypts[,c("Longitude", "Latitude")])
xy <- geom(mypts)[,c("y","x")]
#tps <- Tps(xy, p$spatial)
tps <- Tps(xy, mypts$pkg)
sp <- interpolate(r, tps)
sp <- mask(sp, tza1)
# plot(sp)
# lines(tza1)
```

### Predict maize prices with coodinates only
```{r}
## Predict Maize prices with coodinates only
maize_mypts <- mypts[mypts$Crop == "Maize", ]
rf <- randomForest(pkg ~ Longitude + Latitude , data=maize_mypts)
sp3 <- interpolate(r, rf, xyNames=c("Longitude", "Latitude"))
sp3 <- mask(sp3, tza1)
plot(sp3)
lines(tza1)
```

## Covariates
The covariates used include a mix of crop-specific indicators, temporal variables to capture monthly and yearly effects, geographical coordinates, accessibility measures, climatic conditions, and lagged rainfall to account for delayed effects of weather on crop prices. 
```{r}
ttcity <- rast("C:/H/Tanzania Price data/Datasets/geodata/TRUE/travel/travel_time_to_cities_u5.tif")
ttport <- rast("C:/H/Tanzania Price data/Datasets/geodata/TRUE/travel/travel_time_to_ports_1.tif")
clm    <- rast("C:/H/Tanzania Price data/Datasets/geodata/TRUE/wc2.1_country/TZA_wc2.1_30s_bio.tif")
area   <- rast("C:/H/Tanzania Price data/Datasets/geodata/TRUE/spam/spam2017V2r1_SSA_H_MAIZ_A.tif")
yield  <- rast("C:/H/Tanzania Price data/Datasets/geodata/TRUE/spam/spam2017V2r1_SSA_Y_MAIZ_R.tif")
popd  <- rast("gpw_v4_population_density_rev11_2020_10m.tif")
```

```{r}
names(ttcity) <- c("ttcity_u5") ## travel time cities of 100k or more
names(ttport) <- c("ttport_1") ## travel time to major ports
names(clm) <- gsub("wc2.1_30s_", "", names(clm))
names(area) <- c("MAI_ARE") # SPAM maize area 2010
names(yield)  <- c("MAI_YLD") # SPAM maize yield 2010
names(popd)  <- c("popdens") # GPW4

comment(ttcity) <- "travel time to cities 100k or more"
comment(ttport) <- "travel time to major ports"

comment(popd) <- "population density 2020 (GPW4 @ 10dm)"

comment(clm)[1] <-"BIO1 = Annual Mean Temperature"
comment(clm)[2] <-"BIO2 = Mean Diurnal Range (Mean of monthly (max temp - min temp))"
comment(clm)[3] <-"BIO3 = Isothermality (BIO2/BIO7) (×100)"
comment(clm)[4] <-"BIO4 = Temperature Seasonality (standard deviation ×100)"
comment(clm)[5] <-"BIO5 = Max Temperature of Warmest Month"
comment(clm)[6] <-"BIO6 = Min Temperature of Coldest Month"
comment(clm)[7] <-"BIO7 = Temperature Annual Range (BIO5-BIO6)"
comment(clm)[8] <-"BIO8 = Mean Temperature of Wettest Quarter"
comment(clm)[9] <-"BIO9 = Mean Temperature of Driest Quarter"
comment(clm)[10] <-"BIO10 = Mean Temperature of Warmest Quarter"
comment(clm)[11] <-"BIO11 = Mean Temperature of Coldest Quarter"
comment(clm)[12] <-"BIO12 = Annual Precipitation"
comment(clm)[13] <-"BIO13 = Precipitation of Wettest Month"
comment(clm)[14] <-"BIO14 = Precipitation of Driest Month"
comment(clm)[15] <-"BIO15 = Precipitation Seasonality (Coefficient of Variation)"
comment(clm)[16] <-"BIO16 = Precipitation of Wettest Quarter"
comment(clm)[17] <-"BIO17 = Precipitation of Driest Quarter"
comment(clm)[18] <-"BIO18 = Precipitation of Warmest Quarter"
comment(clm)[19] <-"BIO19 = Precipitation of Coldest Quarter"
```

Harmonize rasters to national boundaries and common resolution
```{r}
ttcity <- resample(ttcity, r)
ttport <- resample(ttport, r)
clm    <- resample(clm, r)
area   <- resample(area, r)
popd   <- resample(popd, r)
freq(is.na(area))
area <- classify(area, cbind(NA,0)) 
yield  <- resample(yield, r)
freq(is.na(yield))
yield <- classify(yield, cbind(NA,0)) 
# check again 
compareGeom(ttcity, ttport, clm, area, yield, popd)
```

```{r}
## Generate Latitude and Longitude grid
latgrd <- longrd <- r
latgrd[] <- yFromCell(latgrd, 1:ncell(latgrd))
longrd[] <- xFromCell(longrd, 1:ncell(longrd))
names(latgrd) <- c("latitude")
names(longrd) <- c("longitude")
```


```{r}
# Convert water polygons to a raster mask
# We will mask out water bodies from our spatiotemporal predictions, as we are only interested in land areas where prices are relevant.
watertz <- terra::vect("Water_Supply_Control-Rivers.shp")
#watertz
water_rast <- rasterize(watertz, r, field = 1, background = 0)
```


Prepare Predictor Stack.
```{r}
rstack <- c(ttcity, ttport, clm, area, yield, popd, latgrd, longrd)
names(rstack)
```


```{r}
# create focal mean to extract from (as alternative to using buffers for extraction, which are not supported in terra)
fm <- focalMat(r, d=0.18, type='circle', fillNA=FALSE)
rstack2 <- focal(rstack, w=fm, fun="mean", na.policy="all", fillvalue=NA, # na.rm=TRUE,
                 expand=TRUE, silent=FALSE) #, filename="", overwrite=FALSE) 
```

```{r}
# extract values to dataset -- use a 20km buffer
# do a focal sum of 20km radius  - this is about 0.18 of a decimal degree... 0.18*112=20.16
fm <- focalMat(r, d=0.18, type='circle', fillNA=FALSE)
rstack2 <- focal(rstack2, w=fm, fun="sum", na.policy="all", fillvalue=NA, na.rm=TRUE,
                 expand=TRUE, silent=FALSE) #, filename="", overwrite=FALSE) 
```

#### Lag rainfall 
```{r}
## Bring in Rainfall Data from Chirps
chirps_path <- "C:/H/Tanzania Price data/chirps_data"

chirps_files <- list.files(chirps_path, pattern = ".tif$", full.names = TRUE)

# Read all CHIRPS data files into a SpatRaster collection
chirps_rasters <- rast(chirps_files)

#crop to Tanzania boundary
Chirps_Tz <- crop(chirps_rasters, tza1)

writeRaster(Chirps_Tz, "Tz_chirps_monthly_croped.tif", overwrite=TRUE)

Tz_chirps_monthly <- terra::rast("Tz_chirps_monthly_croped.tif")
Tz_chirps_monthly

#Replace -9999 with NA
Tz_chirps_monthly <- classify(Tz_chirps_monthly, cbind(-9999,NA))

#extract layer names
layer_names <- names(Tz_chirps_monthly)
# layer_names

# We need to create a sequence of dates from the layer names
# Extract year and month from layer names and convert to Date
dates <- as.Date(paste0(sub("chirps-v2.0\\.", "", layer_names), "-01"), format = "%Y.%m-%d")
# dates

# Assign these dates to the SpatRaster object
time(Tz_chirps_monthly) <- dates

#rename the layers to the formatted dates
names(Tz_chirps_monthly) <- dates

# Check the SpatRaster object
print(Tz_chirps_monthly)
```

```{r}
# do a focal mean of 100km radius - this is about 0.9 of a decimal degree... 0.9009*112=100.9008
# Calculate the focal mean for each layer (month)
fm_r <- focalMat(Tz_chirps_monthly, d=0.9, type='circle', fillNA=FALSE)
Rainfall_focal_sum_100km <- focal(Tz_chirps_monthly, w=fm_r, fun="mean", na.policy="all", fillvalue=NA, na.rm=TRUE,
                                  expand=TRUE, silent=FALSE)
# Check the result
Rainfall_focal_sum_100km
Rainfall <- Rainfall_focal_sum_100km
```

```{r}
#Resample 
Rainfall_res <- resample(Rainfall, r)
Rainfall_res
```


Here we define function that calculates 6 month lag sum of rainfall for each month in our data. The output is raster datsets.
```{r}
calculate_lagged_sum <- function(raster_stack, num_months = 6) {
  # Get the time vector from the raster stack
  time_vector <- time(raster_stack)
  
  # Initialize list to store lagged sum rasters
  lagged_sum_rasters <- vector("list", length(time_vector))
  
  # Loop through each layer in the raster stack
  for (i in seq_along(time_vector)) {
    if (i > num_months) {  # We need at least 'num_months' previous layers to calculate the lagged sum
      # Determine the start and end dates for the lag period
      end_date <- time_vector[i] # Date of the current layer being processed
      start_date <- end_date %m-% months(num_months) #The date num_months before the end_date
      
      # Select the layers that fall within the lag period
      lag_period_layers <- raster_stack[[which(time_vector > start_date & time_vector <= end_date)]]
      
      # Calculate the sum of the selected layers
      if (nlyr(lag_period_layers) == num_months) {
        lagged_sum_rasters[[i]] <- sum(lag_period_layers, na.rm = TRUE)
      } else {
        lagged_sum_rasters[[i]] <- rast(nrow = nrow(raster_stack), ncol = ncol(raster_stack), 
                                        crs = crs(raster_stack), ext = ext(raster_stack), 
                                        vals = NA)  # Use an empty raster with NA values
      }
    } else {
      lagged_sum_rasters[[i]] <- rast(nrow = nrow(raster_stack), ncol = ncol(raster_stack), 
                                      crs = crs(raster_stack), ext = ext(raster_stack), 
                                      vals = NA)  # Use an empty raster with NA values
    }
  }
  
  # Combine the lagged sum rasters into a single raster stack, excluding empty rasters
  lagged_sum_stack <- rast(lagged_sum_rasters)
  
  # Set the names for the layers in the lagged sum stack
  names(lagged_sum_stack) <- names(raster_stack)[!is.na(lagged_sum_rasters)]
  
  return(lagged_sum_stack)
}

# Calculate the 6-month lagged sum for each period in the raster stack
lagged_rainfall_sum <- calculate_lagged_sum(Rainfall_res, num_months = 6)

# Remove the first 6 layers from the raster stack since they are empty
lagged_rainfall_sum_filtered <- lagged_rainfall_sum[[7:nlyr(lagged_rainfall_sum)]]
# check the result
print(lagged_rainfall_sum_filtered)

names(lagged_rainfall_sum_filtered) <- paste0(names(lagged_rainfall_sum_filtered), "_rain.sum.lag")

plot(lagged_rainfall_sum_filtered)
```


```{r}
## We'll have to include lagged_rainfall_sum_filtered in the predictor stack.
rstack
#names(rstack)
```

```{r}
rstack3 <- c(rstack, lagged_rainfall_sum_filtered)
# names(rstack3)
```

```{r}
#Extract to the point dataset
extr1 <- terra::extract(rstack3, mypts, method = "bilinear")
mypts <- cbind(mypts, extr1)
# Remove the ID column from the dataset
mypts <- mypts[, !names(mypts) %in% "ID"]
```


Here we extract sum of lag rainfall for each row under the column rain.sum.lag
```{r}
mypts_df <- as.data.frame(mypts)

# Define the function to obtain sum of lag rainfall from corresponding rasters to mypts under rain.sum.lag column (for each row)
# Each extraction has to match the month and year
get_rain_sum_row <- function(current_date, mypts_row) {
  # Extract the rainfall value for the current date
  rain_sum <- mypts_row[[paste0(current_date, "_rain.sum.lag")]]
  return(rain_sum)
}

# Loop through each row and obtain the rainfall sum for each month and year
for (i in 1:nrow(mypts_df)) {
  # Extract relevant data for the current row
  month <- mypts_df$Month[i]
  year <- mypts_df$Year[i]
  current_date <- paste0(year, "-", sprintf("%02d", month), "-01")  # Format date correctly
  # Pass necessary data to the function
  rain_sum <- get_rain_sum_row(current_date, mypts_df[i, ])
  # Update the rain.sum.lag column
  mypts_df$rain.sum.lag[i] <- rain_sum
}

# Update the SpatVector with the new rain.avg column
mypts$rain.sum.lag <- mypts_df$rain.sum.lag
```

```{r}
# I'll drop the dates with rain.sum.lag from mypts, seems redundant
column_indices <- grep("^202[0-4]-", names(mypts))
mypts <- mypts[, -column_indices]
# names(mypts)
```

```{r}
#define Month as a factor
# mypts$Month <- as.factor(mypts$Month)
# levels(mypts$Month)

# We'll define month as an interger instead.
# Check to make sure Month is interger
# sapply(mypts, class)
```

```{r}
# drop levels that don't exist in Crop field
mypts$Crop <- mypts$Crop[,drop=TRUE]
levels(mypts$Crop)
```

# Linear model 
Coefficient estimates from the linear model provide insights into the relationship between each predictor and the price variable. 
```{r}
# Fit the linear model
# Linear model price Prediction
# Fit the linear model
lm_model <- lm(pkg_real ~ maize + rice + sorghum + bmillet + fmillet + wheat + CPI + beans + Month + Year + ttcity_u5 + ttport_1 + bio_1 + bio_2 + bio_3 + bio_4 + bio_5 + bio_6 +  bio_7 + bio_8 + bio_9 + bio_10 + bio_11 + bio_12 + bio_13 + bio_14 + bio_15 + bio_16 + bio_17 + bio_18 + bio_19 + Longitude + Latitude + 
                 rain.sum.lag,
               data = mypts)
# Extract and print the coefficients
summary(lm_model)
```

# RandomForest 
### Split data the to be used for Training and validation 
Data from May 2021 - Dec 2024 will be used for model training while more recent data from  Jan 2025 - Jan 2026 will be used for Validation.
```{r}
mypts <- as.data.table(mypts)
mypts$Month <- as.integer(as.character(mypts$Month))

training_data <- mypts[mypts$Year %in% c(2021, 2022, 2023, 2024), ]
training_data <- as.data.frame(training_data)
```

```{r}
# Filter the data for validation (Jan 2025 - Jan 2026)
validation_data <- mypts[mypts$Year %in% c(2025, 2026), ]
validation_data <- as.data.frame(validation_data)
```

### Fit The Random Forest Model (1)
#### Random Forest for generating variable of importance
Here, the model is fitted using the randomForest function to build a predictive model for food commodity prices. The goal is to generate  Variable importance scores which will help us understand which variables have the most significant impact on prices, enabling us to interpret the model and possibly simplify it by focusing on the most important predictors.
```{r}
# Create the random forest model
rf1 <- randomForest(pkg_real ~ maize + rice + sorghum + bmillet + fmillet + wheat + beans + potato + Month + Year + CPI + popdens +
                      ttcity_u5 + ttport_1 + 
                      bio_1 + bio_2 + bio_3 + bio_4 + bio_5 + bio_6 + 
                      bio_7 + bio_8 + bio_9 + bio_10 + bio_11 + bio_12 +
                      bio_13 + bio_14 + bio_15 + bio_16 + bio_17 + 
                      bio_18 + bio_19 + MAI_ARE + MAI_YLD + 
                      Longitude + Latitude + 
                      rain.sum.lag,
                    data = training_data,
                    importance=TRUE,na.rm=TRUE)

rf1
```

```{r}
varImpPlot(rf1)
```


```{r}
importance_metrics <- importance(rf1, type=1)  # %IncMSE
impvar <- rownames(importance_metrics)[order(importance_metrics[, 1], decreasing=TRUE)]
# Get the top 20 variables
top_20_vars <- impvar[1:20]
top_20_vars
```

```{r}
node_purity <- importance(rf1, type=2)  # IncNodePurity
# Sort variables by importance (IncNodePurity)
node_purity_sorted <- sort(node_purity[,1], decreasing = TRUE)
# Select the top 20 important variables
top_vars <- names(node_purity_sorted)[1:20]
print(top_vars)
```

```{r}
rf1$importanceSD
```

### Fit The Random Forest Model (2)
#### 1. Tune The Forest
We use the tuneRF function to iteratively search for the optimal mtry value that minimizes the Out-of-Bag (OOB) error. In the random forest model, we will use this mtry value.
```{r}
predictors_pooled <- c("maize", "rice", "sorghum", "bmillet", "fmillet", "wheat", "beans", "potato", "Month", "Year", "CPI", "ttport_1", "ttcity_u5", "popdens", "bio_3", "bio_6", "bio_9", "bio_12", "bio_18", "rain.sum.lag")

trf <- tuneRF(x = training_data[, predictors_pooled], 
              y = training_data$pkg_real, 
              stepFactor = 1.5, improve = 0.01, ntreeTry = 500, plot = FALSE, trace = TRUE)

```

```{r}
best_mtry <- trf[which.min(trf[, 2]), 1]
```


#### 2. Estimate more parsimonious model
Here we fit a more parsimonous model by using the most important predictor variables identified above. This model will be used for spatial-temporal prediction.
```{r}
### Fit The Random Forest Model (2)
rf <- randomForest(pkg_real ~ maize + rice + sorghum + bmillet + fmillet + wheat + beans + potato + Month + Year + CPI + ttport_1 + ttcity_u5 + popdens + bio_3 + bio_6 + bio_9 + bio_12 + bio_18 + rain.sum.lag,
                   data = training_data,
                   mtry = best_mtry,
                   ntree = 1000,
                   nodesize = 5,
                   importance = TRUE,
                   na.action = na.omit)
rf
```

```{r}
# Plotting %IncMSE variable importance plot 
varImpPlot(rf, type = 1)
```

```{r}
(oob <- sqrt(rf$mse[which.min(rf$mse)]))
```

```{r}
partialPlot(rf, as.data.frame(training_data), "rain.sum.lag")
```


## spatial Temporal prediction
These are prediction plots for each food commodities (maize, beans, rice, sorghum, bmillet, fmillet, wheat, potato) with their respective   month of the year 2025.

```{r}
year <- 2025
```

```{r}
# Create a lookup for 2025 CPI values
cpi_2025 <- data.frame(
  Month = 1:12,
  CPI_val = c(117.57, 118.28, 119.27, 119.78, 119.85, 120.18, 
              119.85, 119.77, 119.86, 119.63, 120.01, 121.11))
```

Note: we must set the rain.sum.lag variables for each month we are predicting

### Predicted Maize Prices
```{r}
#Maize
predict_for_maize <- function(month) {
  # Get the actual NCPI for this specific month
  current_cpi <- cpi_2025$CPI_val[cpi_2025$Month == month]
  
  # Load rainfall lag for 2025 (ensure your stack names match this format)
  delta_rain.sum.lag <- rstack3[paste0("2025-", sprintf("%02d", month), "-01_rain.sum.lag")]
  names(delta_rain.sum.lag) <- "rain.sum.lag"
  
  # Prepare the predictor stack
  newstack <- c(rstack, delta_rain.sum.lag)
  watertz_proj <- project(watertz, crs(newstack))
  newstack_masked <- mask(newstack, watertz_proj, inverse = TRUE)
  
  # Define constant data for prediction
  const_maize <- data.frame(
    maize = 1, rice = 0, sorghum = 0, bmillet = 0, fmillet = 0, 
    wheat = 0, beans = 0, potato = 0,
    Month = as.integer(month),
    #Month = factor(month, levels = levels(training_data$Month)),
    Year = 2025,
    CPI = current_cpi  
  )
  
  # Generate prediction
  pred <- predict(newstack_masked, rf, const = const_maize, na.rm = TRUE)
  pred <- mask(pred, tza1)
  
  return(pred)
}

# 3. Create predictions for all months
predictions_maize <- lapply(1:12, predict_for_maize)

# Extract pixel values from predictions_maize
maize_values <- unlist(lapply(predictions_maize, values))
# Get min and max values
min_maize <- min(maize_values, na.rm = TRUE)
min_maize
max_maize <- max(maize_values, na.rm = TRUE)
max_maize

# Define the continuous color palette and reverse it
# color_palette <- rev(terrain.colors(100))
#color_palette <- colorRampPalette(c("blue", "wheat", "red"))(100)
color_palette <- colorRampPalette(c("#081d58", "#41b6c4", "#ffffd9", "#fed976", "#bd0026"))(100)

# Set up plot layout with an extra column for the legend
layout_matrix <- matrix(c(1, 2, 3, 4, 13,
                          5, 6, 7, 8, 13,
                          9, 10, 11, 12, 13), nrow = 3, byrow = TRUE)

# Set up layout
layout(layout_matrix, heights = c(2, 2, 2, 0.6), widths = c(1, 1, 1, 1, 0.6))

break_interval <- 150 

# Create a 3x4 matrix of plots
par(mar = c(0, 0, 0, 0))  # Set margins to 0 for inner plots
for (i in 1:12) {
  plot(predictions_maize[[i]], main = paste(month.name[i], year),
       zlim = c(min_maize, max_maize), col = color_palette, breaks = seq(min_maize, max_maize, by = break_interval), legend = FALSE, axes = TRUE, cex.main = 1.0)
  
  # Add region boundaries
  plot(tza1, add = TRUE, border = "dimgray", lwd = 0.1)
  
  #points(mkt_pts, pch = 19, col = "red", cex = 0.9)
}

# Generate pretty breaks for the legend
legend_breaks <- pretty(c(min_maize, max_maize), n = 4)

# Reset plot layout for the legend
layout(matrix(1))
par(mar = c(5, 4, 2, 4))

# Plot the legend as vertical (default orientation)
image.plot(zlim = c(min_maize, max_maize), legend.only = TRUE,
           col = color_palette, 
           legend.width = 1, legend.shrink = 0.8,
           axis.args = list(at = legend_breaks, labels = legend_breaks, cex.axis = 0.9),
           legend.args = list(text = "", side = 4, line = 3.5, cex = 0.8))

output_dir_Maize <- "C:/Users/LMADAGA/OneDrive - CGIAR/H/Tanzania Price data/Datasets/Pred-plots/Maize"

for (i in 1:length(predictions_maize)) {
  month_str <- sprintf("%02d", i)
  output_filename <- paste0("maize_price_rf_pred_", month_str, ".tif")
  output_path <- file.path(output_dir_Maize, output_filename)
  writeRaster(predictions_maize[[i]], output_path, overwrite = TRUE)
}

```

### Predicted Beans Prices
```{r}
# Beans
# Function to predict beans for a given month
predict_for_beans <- function(month) {
  # Get the actual NCPI for this specific month
  current_cpi <- cpi_2025$CPI_val[cpi_2025$Month == month]
  
  # Load rainfall lag for 2025 (ensure your stack names match this format)
  delta_rain.sum.lag <- rstack3[paste0("2025-", sprintf("%02d", month), "-01_rain.sum.lag")]
  names(delta_rain.sum.lag) <- "rain.sum.lag"
  
  # Prepare the predictor stack
  newstack <- c(rstack, delta_rain.sum.lag)
  watertz_proj <- project(watertz, crs(newstack))
  newstack_masked <- mask(newstack, watertz_proj, inverse = TRUE)
  
  # Define constant data for prediction
  const_maize <- data.frame(
    maize = 0, rice = 0, sorghum = 0, bmillet = 0, fmillet = 0, 
    wheat = 0, beans = 1, potato = 0,
    #Month = factor(month, levels = levels(training_data$Month)),
    Month = as.integer(month),
    Year = 2025,
    CPI = current_cpi  
  )
  
  # Generate prediction
  pred <- predict(newstack_masked, rf, const = const_maize, na.rm = TRUE)
  pred <- mask(pred, tza1)
  
  return(pred)
}
# Create predictions for all months
predictions_beans <- lapply(1:12, predict_for_beans)

# Extract pixel values from predictions_beans
bean_values <- unlist(lapply(predictions_beans, values))
min_bean <- min(bean_values, na.rm = TRUE)
min_bean
max_bean <- max(bean_values, na.rm = TRUE)
max_bean

# Define the continuous color palette and reverse it
#color_palette <- rev(terrain.colors(100))
# color_palette <- colorRampPalette(c("blue", "wheat", "red"))(100)
color_palette <- colorRampPalette(c("#081d58", "#41b6c4", "#ffffd9", "#fed976", "#bd0026"))(100)


# Set up plot layout with an extra row for the legend
layout_matrix <- matrix(c(1, 2, 3, 4, 13,
                          5, 6, 7, 8, 13,
                          9, 10, 11, 12, 13), nrow = 3, byrow = TRUE)

# Set up layout
layout(layout_matrix, heights = c(2, 2, 2, 0.6), widths = c(1, 1, 1, 1, 0.6))


break_interval <- 150 

# Loop through each month to plot beans prices
par(mar = c(0, 0, 0, 0))
for (i in 1:12) {
  plot(predictions_beans[[i]], main = paste(month.name[i], year),
       zlim = c(min_bean, max_bean), col = color_palette, breaks = seq(min_bean, max_bean, by = break_interval), legend = FALSE, axes = TRUE, cex.main = 1.0)
  
  plot(tza1, add = TRUE, border = "dimgray", lwd = 0.1)
  
  points(training_data, pch = 20, col = "red", cex = 0.5)
}

# Generate pretty breaks for the legend
legend_breaks <- pretty(c(min_bean, max_bean), n = 4)

# Reset plot layout to 1x1 for the legend
layout(matrix(1))
par(mar = c(5, 4, 2, 4)) 

# Plot the legend in the bottom row, centered horizontally
image.plot(zlim = c(min_bean, max_bean), legend.only = TRUE, 
           col = color_palette,
           legend.width = 0.7, legend.shrink = 0.9, 
           axis.args = list(at = legend_breaks, labels = legend_breaks, cex.axis = 0.9), 
           legend.args = list(text = "", side = 2, line = 2, cex = 0.9))


output_dir_beans <- "C:/Users/LMADAGA/OneDrive - CGIAR/H/Tanzania Price data/Datasets/Pred-plots/beans"

for (i in 1:length(predictions_beans)) {
  month_str <- sprintf("%02d", i)
  output_filename <- paste0("beans_price_rf_pred_", month_str, ".tif")
  output_path <- file.path(output_dir_beans, output_filename)
  writeRaster(predictions_beans[[i]], output_path, overwrite = TRUE)
}

```

### Predicted Rice Prices
```{r}
# Rice
# Function to predict rice prices for a given month
predict_for_rice <- function(month) {
  # Get the actual NCPI for this specific month
  current_cpi <- cpi_2025$CPI_val[cpi_2025$Month == month]
  
  # Load rainfall lag for 2025 (ensure your stack names match this format)
  delta_rain.sum.lag <- rstack3[paste0("2025-", sprintf("%02d", month), "-01_rain.sum.lag")]
  names(delta_rain.sum.lag) <- "rain.sum.lag"
  
  # Prepare the predictor stack
  newstack <- c(rstack, delta_rain.sum.lag)
  watertz_proj <- project(watertz, crs(newstack))
  newstack_masked <- mask(newstack, watertz_proj, inverse = TRUE)
  
  # Define constant data for prediction
  const_maize <- data.frame(
    maize = 0, rice = 1, sorghum = 0, bmillet = 0, fmillet = 0, 
    wheat = 0, beans = 0, potato = 0,
    # Month = factor(month, levels = levels(training_data$Month)),
    Month = as.integer(month),
    Year = 2025,
    CPI = current_cpi  
  )
  
  # Generate prediction
  pred <- predict(newstack_masked, rf, const = const_maize, na.rm = TRUE)
  pred <- mask(pred, tza1)
  
  return(pred)
}

# Create predictions for all months
predictions_rice <- lapply(1:12, predict_for_rice)

# Extract pixel values from predictions_rice
rice_values <- unlist(lapply(predictions_rice, values))
min_rice <- min(rice_values, na.rm = TRUE)
min_rice
max_rice <- max(rice_values, na.rm = TRUE)
max_rice

# Define the continuous color palette and reverse it
# color_palette <- rev(terrain.colors(100))
#color_palette <- colorRampPalette(c("blue", "wheat", "red"))(100)
color_palette <- colorRampPalette(c("#081d58", "#41b6c4", "#ffffd9", "#fed976", "#bd0026"))(100)

layout_matrix <- matrix(c(1, 2, 3, 4, 13,
                          5, 6, 7, 8, 13,
                          9, 10, 11, 12, 13), nrow = 3, byrow = TRUE)

# Set up layout
layout(layout_matrix, heights = c(2, 2, 2, 0.6), widths = c(1, 1, 1, 1, 0.6))


break_interval <- 150 

# Loop through each month to plot rice prices
par(mar = c(0, 0, 0, 0))
for (i in 1:12) {
  plot(predictions_rice[[i]], main = paste(month.name[i], year),
       zlim = c(min_rice, max_rice), col = color_palette, breaks = seq(min_rice, max_rice, by = break_interval), legend = FALSE, axes = TRUE, cex.main = 1.0)
  
  plot(tza1, add = TRUE, border = "dimgray", lwd = 0.1)
  
  points(training_data, pch = 20, col = "red", cex = 0.5)
}

# Generate pretty breaks for the legend
legend_breaks <- pretty(c(min_rice, max_rice), n = 5)

# Reset plot layout to 1x1 for the legend
layout(matrix(1))
par(mar = c(5, 4, 2, 4))  

# Plot the legend in the bottom row, centered horizontally
image.plot(zlim = c(min_rice, max_rice), legend.only = TRUE, 
           col = color_palette,
           legend.width = 0.7, legend.shrink = 0.9, 
           axis.args = list(at = legend_breaks, labels = legend_breaks, cex.axis = 0.9), 
           legend.args = list(text = "", side = 1, line = 2, cex = 0.9))

output_dir_rice <- "C:/Users/LMADAGA/OneDrive - CGIAR/H/Tanzania Price data/Datasets/Pred-plots/rice"

for (i in 1:length(predictions_rice)) {
  month_str <- sprintf("%02d", i)
  output_filename <- paste0("rice_price_rf_pred_", month_str, ".tif")
  output_path <- file.path(output_dir_rice, output_filename)
  writeRaster(predictions_rice[[i]], output_path, overwrite = TRUE)
}

```

### Predicted Sorghum Prices
```{r}
# sorghum
# Function to predict sorghum prices for a given month
predict_for_sorghum <- function(month) {
  # Get the actual NCPI for this specific month
  current_cpi <- cpi_2025$CPI_val[cpi_2025$Month == month]
  
  # Load rainfall lag for 2025 (ensure your stack names match this format)
  delta_rain.sum.lag <- rstack3[paste0("2025-", sprintf("%02d", month), "-01_rain.sum.lag")]
  names(delta_rain.sum.lag) <- "rain.sum.lag"
  
  # Prepare the predictor stack
  newstack <- c(rstack, delta_rain.sum.lag)
  watertz_proj <- project(watertz, crs(newstack))
  newstack_masked <- mask(newstack, watertz_proj, inverse = TRUE)
  
  # Define constant data for prediction
  const_maize <- data.frame(
    maize = 0, rice = 0, sorghum = 1, bmillet = 0, fmillet = 0, 
    wheat = 0, beans = 0, potato = 0,
    # Month = factor(month, levels = levels(training_data$Month)),
    Month = as.integer(month),
    Year = 2025,
    CPI = current_cpi  
  )
  
  # Generate prediction
  pred <- predict(newstack_masked, rf, const = const_maize, na.rm = TRUE)
  pred <- mask(pred, tza1)
  
  return(pred)
}

# Create predictions for all months
predictions_sorghum <- lapply(1:12, predict_for_sorghum)

# Extract pixel values from predictions_sorghum
sorghum_values <- unlist(lapply(predictions_sorghum, values))
min_sorghum <- min(sorghum_values, na.rm = TRUE)
min_sorghum
max_sorghum <- max(sorghum_values, na.rm = TRUE)
max_sorghum

# Define the continuous color palette and reverse it
#color_palette <- colorRampPalette(c("blue", "wheat", "red"))(100)
color_palette <- colorRampPalette(c("#081d58", "#41b6c4", "#ffffd9", "#fed976", "#bd0026"))(100)

# Set up plot layout with an extra row for the legend
layout_matrix <- matrix(c(1, 2, 3, 4, 13,
                          5, 6, 7, 8, 13,
                          9, 10, 11, 12, 13), nrow = 3, byrow = TRUE)

# Set up layout
layout(layout_matrix, heights = c(2, 2, 2, 0.6), widths = c(1, 1, 1, 1, 0.6))


break_interval <- 250 

# Loop through each month to plot sorghum prices
par(mar = c(0, 0, 0, 0))
for (i in 1:12) {
  plot(predictions_sorghum[[i]], main = paste(month.name[i], year),
       zlim = c(min_sorghum, max_sorghum), col = color_palette, breaks = seq(min_sorghum, max_sorghum, by = break_interval), legend = FALSE, axes = TRUE, cex.main = 1.0)
  
  plot(tza1, add = TRUE, border = "dimgray", lwd = 0.1)
  
  points(training_data, pch = 20, col = "red", cex = 0.5)
}

# Generate pretty breaks for the legend
legend_breaks <- pretty(c(min_sorghum, max_sorghum), n = 5)

# Reset plot layout to 1x1 for the legend
layout(matrix(1))
par(mar = c(5, 4, 2, 4))  

# Plot the legend in the bottom row, centered horizontally
image.plot(zlim = c(min_sorghum, max_sorghum), legend.only = TRUE, 
           col = color_palette,
           legend.width = 0.7, legend.shrink = 0.9, 
           axis.args = list(at = legend_breaks, labels = legend_breaks, cex.axis = 0.9), 
           legend.args = list(text = "", side = 1, line = 2, cex = 0.9))

output_dir_sorghum <- "C:/Users/LMADAGA/OneDrive - CGIAR/H/Tanzania Price data/Datasets/Pred-plots/sorghum"

for (i in 1:length(predictions_sorghum)) {
  month_str <- sprintf("%02d", i)
  output_filename <- paste0("sorghum_price_rf_pred_", month_str, ".tif")
  output_path <- file.path(output_dir_sorghum, output_filename)
  writeRaster(predictions_sorghum[[i]], output_path, overwrite = TRUE)
}

```

### Predicted Bulrush Millet Prices
```{r}
# bmillet
# Function to predict bmillet prices for a given month
predict_for_bmillet <- function(month) {
  # Get the actual NCPI for this specific month
  current_cpi <- cpi_2025$CPI_val[cpi_2025$Month == month]
  
  # Load rainfall lag for 2025
  delta_rain.sum.lag <- rstack3[paste0("2025-", sprintf("%02d", month), "-01_rain.sum.lag")]
  names(delta_rain.sum.lag) <- "rain.sum.lag"
  
  # Prepare the predictor stack
  newstack <- c(rstack, delta_rain.sum.lag)
  watertz_proj <- project(watertz, crs(newstack))
  newstack_masked <- mask(newstack, watertz_proj, inverse = TRUE)
  
  # Define constant data for prediction
  const_maize <- data.frame(
    maize = 0, rice = 0, sorghum = 0, bmillet = 1, fmillet = 0, 
    wheat = 0, beans = 0, potato = 0,
    # Month = factor(month, levels = levels(training_data$Month)),
    Month = as.integer(month),
    Year = 2025,
    CPI = current_cpi  
  )
  
  # Generate prediction
  pred <- predict(newstack_masked, rf, const = const_maize, na.rm = TRUE)
  pred <- mask(pred, tza1)
  
  return(pred)
}

# Create predictions for all months
predictions_bmillet <- lapply(1:12, predict_for_bmillet)

# Extract pixel values from predictions_bmillet
bmillet_values <- unlist(lapply(predictions_bmillet, values))
min_bmillet <- min(bmillet_values, na.rm = TRUE)
min_bmillet
max_bmillet <- max(bmillet_values, na.rm = TRUE)
max_bmillet

# Define the continuous color palette and reverse it
# color_palette <- rev(terrain.colors(100))
#color_palette <- colorRampPalette(c("blue", "wheat", "red"))(100)
color_palette <- colorRampPalette(c("#081d58", "#41b6c4", "#ffffd9", "#fed976", "#bd0026"))(100)

# Set up plot layout with an extra row for the legend
layout_matrix <- matrix(c(1, 2, 3, 4, 13,
                          5, 6, 7, 8, 13,
                          9, 10, 11, 12, 13), nrow = 3, byrow = TRUE)

# Set up layout
layout(layout_matrix, heights = c(2, 2, 2, 0.6), widths = c(1, 1, 1, 1, 0.6))

# Loop through each month to plot bmillet prices
break_interval <- 200 
par(mar = c(0, 0, 0, 0)) 
for (i in 1:12) {
  plot(predictions_bmillet[[i]], main = paste(month.name[i], year),
       zlim = c(min_bmillet, max_bmillet), col = color_palette, 
       breaks = seq(min_bmillet, max_bmillet, by = break_interval), 
       legend = FALSE, axes = TRUE, cex.main = 1.0)
  
  plot(tza1, add = TRUE, border = "dimgray", lwd = 0.1)
  
  #points(training_data, pch = 20, col = "red", cex = 0.5)
}

# Generate pretty breaks for the legend
legend_breaks <- pretty(c(min_bmillet, max_bmillet), n = 5) 

# Reset plot layout to 1x1 for the legend
layout(matrix(1))
par(mar = c(5, 4, 2, 4))  

# Plot the legend in the bottom row, centered horizontally
image.plot(zlim = c(min_bmillet, max_bmillet), legend.only = TRUE, 
           col = color_palette, 
           legend.width = 0.7, legend.shrink = 0.9, 
           axis.args = list(at = legend_breaks, labels = legend_breaks, cex.axis = 0.9), 
           legend.args = list(text = "", side = 1, line = 2, cex = 0.9))

output_dir_bmillet <- "C:/Users/LMADAGA/OneDrive - CGIAR/H/Tanzania Price data/Datasets/Pred-plots/bmillet"

for (i in 1:length(predictions_bmillet)) {
  month_str <- sprintf("%02d", i)
  output_filename <- paste0("bmillet_price_rf_pred_", month_str, ".tif")
  output_path <- file.path(output_dir_bmillet, output_filename)
  writeRaster(predictions_bmillet[[i]], output_path, overwrite = TRUE)
}

```

### Predicted Finger Millet Prices
```{r}
#fmillet
# Function to predict fmillet prices for a given month
predict_for_fmillet <- function(month) {
  # Get the actual NCPI for this specific month
  current_cpi <- cpi_2025$CPI_val[cpi_2025$Month == month]
  
  # Load rainfall lag for 2025
  delta_rain.sum.lag <- rstack3[paste0("2025-", sprintf("%02d", month), "-01_rain.sum.lag")]
  names(delta_rain.sum.lag) <- "rain.sum.lag"
  
  # Prepare the predictor stack
  newstack <- c(rstack, delta_rain.sum.lag)
  watertz_proj <- project(watertz, crs(newstack))
  newstack_masked <- mask(newstack, watertz_proj, inverse = TRUE)
  
  # Define constant data for prediction
  const_maize <- data.frame(
    maize = 0, rice = 0, sorghum = 0, bmillet = 0, fmillet = 1, 
    wheat = 0, beans = 0, potato = 0,
    #Month = factor(month, levels = levels(training_data$Month)),
    Month = as.integer(month),
    Year = 2025,
    CPI = current_cpi  
  )
  
  # Generate prediction
  pred <- predict(newstack_masked, rf, const = const_maize, na.rm = TRUE)
  pred <- mask(pred, tza1)
  
  return(pred)
}

# Create predictions for all months
predictions_fmillet <- lapply(1:12, predict_for_fmillet)

# Extract pixel values from predictions_fmillet
fmillet_values <- unlist(lapply(predictions_fmillet, values))
min_fmillet <- min(fmillet_values, na.rm = TRUE)
max_fmillet <- max(fmillet_values, na.rm = TRUE)

# Define the continuous color palette and reverse it
#color_palette <- rev(terrain.colors(100))
#color_palette <- colorRampPalette(c("blue", "wheat", "red"))(100)
color_palette <- colorRampPalette(c("#081d58", "#41b6c4", "#ffffd9", "#fed976", "#bd0026"))(100)

# Set up plot layout with an extra row for the legend
layout_matrix <- matrix(c(1, 2, 3, 4, 13,
                          5, 6, 7, 8, 13,
                          9, 10, 11, 12, 13), nrow = 3, byrow = TRUE)

# Set up layout
layout(layout_matrix, heights = c(2, 2, 2, 0.6), widths = c(1, 1, 1, 1, 0.6))

# Loop through each month to plot fmillet prices
break_interval <- 100 
par(mar = c(0, 0, 0, 0))
for (i in 1:12) {
  plot(predictions_fmillet[[i]], main = paste(month.name[i], year),
       zlim = c(min_fmillet, max_fmillet), col = color_palette, 
       breaks = seq(min_fmillet, max_fmillet, by = break_interval), 
       legend = FALSE, axes = TRUE, cex.main = 1.0)
  
  plot(tza1, add = TRUE, border = "dimgray", lwd = 0.1)
  
  points(training_data, pch = 20, col = "red", cex = 0.5)
}

# Generate pretty breaks for the legend
legend_breaks <- pretty(c(min_fmillet, max_fmillet), n = 3)  

# Reset plot layout to 1x1 for the legend
layout(matrix(1))
par(mar = c(5, 4, 2, 4))  

# Plot the legend in the bottom row, centered horizontally
image.plot(zlim = c(min_fmillet, max_fmillet), legend.only = TRUE, 
           col = color_palette,
           legend.width = 0.7, legend.shrink = 0.9, 
           axis.args = list(at = legend_breaks, labels = legend_breaks, cex.axis = 0.9), 
           legend.args = list(text = "", side = 1, line = 2, cex = 0.9))

output_dir_fmillet <- "C:/Users/LMADAGA/OneDrive - CGIAR/H/Tanzania Price data/Datasets/Pred-plots/fmillet"

for (i in 1:length(predictions_fmillet)) {
  month_str <- sprintf("%02d", i)
  output_filename <- paste0("fmillet_price_rf_pred_", month_str, ".tif")
  output_path <- file.path(output_dir_fmillet, output_filename)
  writeRaster(predictions_fmillet[[i]], output_path, overwrite = TRUE)
}

```

### Predicted Wheat Prices
```{r}
#Wheat
# Function to predict wheat prices for a given month
predict_for_wheat <- function(month) {
  # Get the actual NCPI for this specific month
  current_cpi <- cpi_2025$CPI_val[cpi_2025$Month == month]
  
  # Load rainfall lag for 2025
  delta_rain.sum.lag <- rstack3[paste0("2025-", sprintf("%02d", month), "-01_rain.sum.lag")]
  names(delta_rain.sum.lag) <- "rain.sum.lag"
  
  # Prepare the predictor stack
  newstack <- c(rstack, delta_rain.sum.lag)
  watertz_proj <- project(watertz, crs(newstack))
  newstack_masked <- mask(newstack, watertz_proj, inverse = TRUE)
  
  # Define constant data for prediction
  const_maize <- data.frame(
    maize = 0, rice = 0, sorghum = 0, bmillet = 0, fmillet = 0, 
    wheat = 1, beans = 0, potato = 0,
    #Month = factor(month, levels = levels(training_data$Month)),
    Month = as.integer(month),
    Year = 2025,
    CPI = current_cpi  
  )
  
  # Generate prediction
  pred <- predict(newstack_masked, rf, const = const_maize, na.rm = TRUE)
  pred <- mask(pred, tza1)
  
  return(pred)
}

# Create predictions for all months
predictions_wheat <- lapply(1:12, predict_for_wheat)
# Extract pixel values from predictions_wheat
wheat_values <- unlist(lapply(predictions_wheat, values))
min_wheat <- min(wheat_values, na.rm = TRUE)
min_wheat
max_wheat <- max(wheat_values, na.rm = TRUE)
max_wheat

# Define the continuous color palette and reverse it
# color_palette <- rev(terrain.colors(100))
#color_palette <- colorRampPalette(c("blue", "wheat", "red"))(100)
color_palette <- colorRampPalette(c("#081d58", "#41b6c4", "#ffffd9", "#fed976", "#bd0026"))(100)

# Set up plot layout with an extra row for the legend
layout_matrix <- matrix(c(1, 2, 3, 4, 13,
                          5, 6, 7, 8, 13,
                          9, 10, 11, 12, 13), nrow = 3, byrow = TRUE)

# Set up layout
layout(layout_matrix, heights = c(2, 2, 2, 0.6), widths = c(1, 1, 1, 1, 0.6))


# Define the break interval for both plot and legend
break_interval <- 300

# Loop through each month to plot wheat prices
par(mar = c(0, 0, 0, 0))
for (i in 1:12) {
  plot(predictions_wheat[[i]], main = paste(month.name[i], year),
       zlim = c(min_wheat, max_wheat), col = color_palette, 
       breaks = seq(min_wheat, max_wheat, by = break_interval), 
       legend = FALSE, axes = TRUE, cex.main = 1.0)
  
  plot(tza1, add = TRUE, border = "dimgray", lwd = 0.1)
  
  points(training_data, pch = 20, col = "red", cex = 0.5)
}

# Generate pretty breaks for the legend
legend_breaks <- pretty(c(min_wheat, max_wheat), n = 5)

# Reset plot layout to 1x1 for the legend
layout(matrix(1))
par(mar = c(5, 4, 2, 4))    

# Plot the legend in the bottom row, centered horizontally
image.plot(zlim = c(min_wheat, max_wheat), legend.only = TRUE, 
           col = color_palette,
           legend.width = 0.7, legend.shrink = 0.9, 
           axis.args = list(at = legend_breaks, labels = legend_breaks, cex.axis = 0.9), 
           legend.args = list(text = "", side = 1, line = 2, cex = 0.9))

output_dir_wheat <- "C:/Users/LMADAGA/OneDrive - CGIAR/H/Tanzania Price data/Datasets/Pred-plots/wheat"

#save
for (i in 1:length(predictions_wheat)) {
  month_str <- sprintf("%02d", i)
  output_filename <- paste0("wheat_price_rf_pred_", month_str, ".tif")
  output_path <- file.path(output_dir_wheat, output_filename)
  writeRaster(predictions_wheat[[i]], output_path, overwrite = TRUE)
}

```

## Predicted potato prices
```{r}
#potatoes
# Function to predict potato prices for a given month
predict_for_potato <- function(month) {
  # Get the actual NCPI for this specific month
  current_cpi <- cpi_2025$CPI_val[cpi_2025$Month == month]
  
  # Load rainfall lag for 2025
  delta_rain.sum.lag <- rstack3[paste0("2025-", sprintf("%02d", month), "-01_rain.sum.lag")]
  names(delta_rain.sum.lag) <- "rain.sum.lag"
  
  # Prepare the predictor stack
  newstack <- c(rstack, delta_rain.sum.lag)
  watertz_proj <- project(watertz, crs(newstack))
  newstack_masked <- mask(newstack, watertz_proj, inverse = TRUE)
  
  # Define constant data for prediction
  const_maize <- data.frame(
    maize = 0, rice = 0, sorghum = 0, bmillet = 0, fmillet = 0, 
    wheat = 0, beans = 0, potato = 1,
    # Month = factor(month, levels = levels(training_data$Month)),
    Month = as.integer(month),
    Year = 2025,
    CPI = current_cpi  
  )
  
  # Generate prediction
  pred <- predict(newstack_masked, rf, const = const_maize, na.rm = TRUE)
  pred <- mask(pred, tza1)
  
  return(pred)
}


# Create predictions for all months
predictions_potato <- lapply(1:12, predict_for_potato)

# Extract pixel values from predictions_potato
potato_values <- unlist(lapply(predictions_potato, values))
min_potato <- min(potato_values, na.rm = TRUE)
min_potato
max_potato <- max(potato_values, na.rm = TRUE)
max_potato

# Define the continuous color palette and reverse it
# color_palette <- rev(terrain.colors(100))
#color_palette <- colorRampPalette(c("blue", "wheat", "red"))(100)
color_palette <- colorRampPalette(c("#081d58", "#41b6c4", "#ffffd9", "#fed976", "#bd0026"))(100)


# Set up plot layout with an extra row for the legend
layout_matrix <- matrix(c(1, 2, 3, 4, 13,
                          5, 6, 7, 8, 13,
                          9, 10, 11, 12, 13), nrow = 3, byrow = TRUE)

# Set up layout
layout(layout_matrix, heights = c(2, 2, 2, 0.6), widths = c(1, 1, 1, 1, 0.6))


# Loop through each month to plot potato prices
break_interval <- 150
par(mar = c(0, 0, 0, 0))
for (i in 1:12) {
  plot(predictions_potato[[i]],  main = paste(month.name[i], year),
       zlim = c(min_potato, max_potato), col = color_palette, 
       breaks = seq(min_potato, max_potato, by = break_interval), 
       legend = FALSE, axes = TRUE,cex.main = 1.0)
  
  plot(tza1, add = TRUE, border = "dimgray", lwd = 0.1)
  
  points(training_data, pch = 20, col = "red", cex = 0.5)
}

# Generate pretty breaks for the legend
legend_breaks <- pretty(c(min_potato, max_potato), n = 5) 

# Reset plot layout to 1x1 for the legend
layout(matrix(1))
par(mar = c(5, 4, 2, 4)) 

# Plot the legend in the bottom row, centered horizontally
image.plot(zlim = c(min_potato, max_potato), legend.only = TRUE, 
           col = color_palette,
           legend.width = 0.7, legend.shrink = 0.9, 
           axis.args = list(at = legend_breaks, labels = legend_breaks, cex.axis = 0.9), 
           legend.args = list(text = "", side = 1, line = 2, cex = 0.9))

output_dir_potato <- "C:/Users/LMADAGA/OneDrive - CGIAR/H/Tanzania Price data/Datasets/Pred-plots/potato"

#save
for (i in 1:length(predictions_potato)) {
  month_str <- sprintf("%02d", i)
  output_filename <- paste0("potato_price_rf_pred_", month_str, ".tif")
  output_path <- file.path(output_dir_potato, output_filename)
  writeRaster(predictions_potato[[i]], output_path, overwrite = TRUE)
}
```


# Comparing Pooled Vs Crop Specific Price Predictions Using different cross validation methods
## 1. Temporal split validation - Standard RF

```{r}
# Define the Split
# Training: 2021-2024 | Validation: 2025-2026
# sapply(mypts, class)
training_data <- mypts %>% filter(Year <= 2024) %>% as.data.frame()
validation_data <- mypts %>% filter(Year >= 2025) %>% as.data.frame()

# Hyperparameter Tuning for the Pooled Model

# Note: tuneRF can be slow; we use only the required columns to speed it up
predictors_pooled <- c("maize", "rice", "sorghum", "bmillet", "fmillet", "wheat", "beans", "potato", "Month", "Year", "CPI", "ttport_1", "ttcity_u5", "popdens", 
 "bio_3", "bio_6", "bio_9", "bio_12", "bio_18", "rain.sum.lag")

trf <- tuneRF(x = training_data[, predictors_pooled], 
              y = training_data$pkg_real, 
              stepFactor = 1.5, improve = 0.01, ntreeTry = 500, plot = FALSE, trace = TRUE)

best_mtry <- trf[which.min(trf[, 2]), 1]


# Fit the Pooled Temporal RF
rf_pooled_temp <- randomForest(
  formula = pkg_real ~ ., 
  data = training_data[, c("pkg_real", predictors_pooled)],
  mtry = best_mtry, 
  ntree = 1000, 
  nodesize = 5, 
  importance = TRUE)

rf_pooled_temp

# Evaluation Function (Pooled vs. Crop-Specific for standard RF)
evaluate_temporal_models <- function(target_crop) {
  cat("Processing:", target_crop, "\n")
  
  # Filter crop data
  train_c <- training_data %>% filter(Crop == target_crop)
  valid_c <- validation_data %>% filter(Crop == target_crop)
  
  # Skip if no data for this crop in validation period
  if(nrow(valid_c) == 0) return(NULL)
  
  # POOLED MODEL PERFORMANCE
  # Predicting 2025/26 using the model trained on all crops 2021-2024
  pooled_pred <- predict(rf_pooled_temp, newdata = valid_c)
  
  # CROP-SPECIFIC MODEL
  # Predictors for single crop (exclude all other crop dummies)
  preds_spec <- c("Month", "Year", "CPI", "ttport_1", "ttcity_u5", "popdens", 
                  "bio_3", "bio_6", "bio_9", "bio_12", "bio_18", "rain.sum.lag")
  
  rf_specific <- randomForest(
    pkg_real ~ ., 
    data = train_c[, c("pkg_real", preds_spec)],
    ntree = 1000, 
    nodesize = 5)
  
  spec_pred <- predict(rf_specific, newdata = valid_c)
  
  # PERFOMANCE METRICS 
  actual <- valid_c$pkg_real
  
  res <- data.frame(
    Crop = target_crop,
    RMSE_Pooled = round(sqrt(mean((actual - pooled_pred)^2, na.rm=TRUE)), 2),
    R2_Pooled   = round(cor(actual, pooled_pred)^2, 2), 
    RMSE_Specific = round(sqrt(mean((actual - spec_pred)^2, na.rm=TRUE)), 2),
    R2_Specific   = round(cor(actual, spec_pred)^2, 2)
  )
  return(res)
}

# Run Comparison
all_crops <- unique(mypts$Crop)
set.seed(1983)
comparison_results_ts <- do.call(rbind, lapply(all_crops, evaluate_temporal_models))

print(comparison_results_ts)
```


```{r}
# Clean up the results directly without pivot_wider
comparison_df_wide <- comparison_results_ts %>%
  rename(
    R_squared_Pooled = R2_Pooled,
    R_squared_Crop_Specific = R2_Specific,
    RMSE_Crop_Specific = RMSE_Specific)

```

```{r}
knitr::kable(comparison_df_wide, caption = "Model Comparison: R² and RMSE (Temporal Validation 2025-2026)")
```

```{r}
write.csv(comparison_df_wide, "Temporal-split-results_std_RF.csv", row.names = FALSE)
```

# 2. Autoregressive Random Forest
 This approach involves incorporating lagged Prices as predictors to capture temporal dependencies in the data. 
 
```{r}
mypts <- as.data.frame(mypts) %>%
  arrange(Crop, Market, Year, Month) %>%
  group_by(Crop, Market) %>%
  mutate(pkg_lag1 = lag(pkg_real, 1),
         pkg_lag2 = lag(pkg_real, 2),
         pkg_lag3 = lag(pkg_real, 3)) %>%
  ungroup() %>%
  na.omit() 
```

```{r}
# Training: 2021-2024 | Validation: 2025-2026
training_data <- mypts %>% filter(Year <= 2024)
validation_data <- mypts %>% filter(Year >= 2025)
```


```{r}
# Hyperparameter Tuning for Pooled AR 
predictors_ar <- c("maize", "rice", "sorghum", "bmillet", "fmillet", "wheat", "beans", "potato", "Month", "Year", "CPI", "ttport_1", "ttcity_u5", "popdens", 
 "bio_3", "bio_6", "bio_9", "bio_12", "bio_18", "rain.sum.lag", "pkg_lag1", "pkg_lag2", "pkg_lag3")

cat("Tuning Pooled AR Model...\n")
trf_ar <- tuneRF(x = training_data[, predictors_ar], y = training_data$pkg_real, 
                 stepFactor = 1.5, improve = 0.01, ntreeTry = 500, plot = FALSE, trace = FALSE)
best_mtry_ar <- trf_ar[which.min(trf_ar[, 2]), 1]

# Fit Final Pooled AR
rf_pooled_ar <- randomForest(pkg_real ~ ., data = training_data[, c("pkg_real", predictors_ar)], mtry = best_mtry_ar, ntree = 1000, importance = TRUE)

rf_pooled_ar

varImpPlot(rf_pooled_ar)
```

```{r}
# Evaluation Function (Pooled AR vs. Crop-Specific AR)
evaluate_ar_models <- function(target_crop) {
  cat("AR Evaluation for:", target_crop, "\n")
  
  train_c <- training_data %>% filter(Crop == target_crop)
  valid_c <- validation_data %>% filter(Crop == target_crop)
  
  if(nrow(valid_c) == 0) return(NULL)
  
  # Pooled AR Prediction
  pred_pooled <- predict(rf_pooled_ar, newdata = valid_c)
  
  # Crop-Specific AR Model
  # Exclude crop indicators for specific model
  preds_spec <- c("Month", "Year", "CPI", "ttport_1", "ttcity_u5", "popdens", 
                  "bio_3", "bio_6", "bio_9", "bio_12", "bio_18", "rain.sum.lag",
                  "pkg_lag1", "pkg_lag2", "pkg_lag3")
  
  # Specific mtry tuning for fairness
  trf_spec <- tuneRF(x = train_c[, preds_spec], y = train_c$pkg_real, 
                     stepFactor = 1.5, improve = 0.01, ntreeTry = 200, plot = FALSE, trace = FALSE)
  
mtry_spec <- trf_spec[which.min(trf_spec[, 2]), 1]
  
rf_spec_ar <- randomForest(pkg_real ~ ., data = train_c[, c("pkg_real", preds_spec)], mtry = mtry_spec, ntree = 1000)
  
  pred_spec <- predict(rf_spec_ar, newdata = valid_c)
  
  # Metrics
  actual <- valid_c$pkg_real
  return(data.frame(
    Crop = target_crop,
    RMSE_Pooled_AR = round(sqrt(mean((actual - pred_pooled)^2, na.rm=TRUE)), 2),
    R2_Pooled_AR   = round(cor(actual, pred_pooled)^2, 2),
    RMSE_Spec_AR   = round(sqrt(mean((actual - pred_spec)^2, na.rm=TRUE)), 2),
    R2_Spec_AR     = round(cor(actual, pred_spec)^2, 2)
  ))
}

# Execute and Compare 
set.seed(1983)
comparison_ar_final <- do.call(rbind, lapply(unique(mypts$Crop), evaluate_ar_models))
print(comparison_ar_final)

```


```{r}
comparison_df_wide_ar <- comparison_ar_final %>%
  rename(
    RMSE_Pooled = RMSE_Pooled_AR,
    R_squared_Pooled = R2_Pooled_AR,
    RMSE_Crop_Specific = RMSE_Spec_AR,
    R_squared_Crop_Specific = R2_Spec_AR)

```

```{r}
knitr::kable(comparison_df_wide_ar, digits = 2, caption = "Table 2: Autoregressive (AR) Model Comparison - Pooled vs. Crop-Specific")
```

```{r}
write.csv(comparison_df_wide_ar, "Temporal-split-results_AR.csv", row.names = FALSE)
```


## 3. Random split validation
Here the dataset is split once (70:30). The model is trained on one part of the data (70%) and evaluated on the remaining set of the data (30%).
```{r}
# Prepare Pooled Data to Split
mypts <- as.data.frame(mypts)
# sapply(mypts, class)
rows <- sample(x = 1:nrow(mypts), size = 0.70 * nrow(mypts))
train_pooled <- mypts[rows, ]
test_pooled  <- mypts[-rows, ]


# Hyperparameter Tuning for Pooled Model
# Define predictors for pooled (including crop dummies)
predictors_pooled <- c("maize", "rice", "sorghum", "bmillet", "fmillet", "wheat", "beans", "potato", "Month", "Year", "CPI", "ttport_1", "ttcity_u5", "popdens", 
"bio_3", "bio_6", "bio_9", "bio_12", "bio_18", "rain.sum.lag")

x_vars_p <- train_pooled[, predictors_pooled]
y_var_p  <- train_pooled$pkg_real

cat("Tuning Pooled Model...\n")

trf_p <- tuneRF(x = x_vars_p, y = y_var_p, stepFactor = 1.5, improve = 0.01, ntreeTry = 500, plot = FALSE, trace = FALSE)

best_mtry_p <- trf_p[which.min(trf_p[, 2]), 1]

# Fit Final Pooled Model
RF_pooled <- randomForest(pkg_real ~ ., data = train_pooled[, c("pkg_real", predictors_pooled)], mtry = best_mtry_p, ntree = 1000, nodesize = 5, importance = TRUE)

RF_pooled

```

```{r}
# Evaluation Function: Pooled vs. Crop-Specific
evaluate_models <- function(target_crop) {
  cat("Evaluating Crop:", target_crop, "\n")
  
  # Filter data for this specific crop
  crop_data <- mypts[mypts$Crop == target_crop, ]
  
  # Split this crop's data 70/30
  set.seed(1983) 
  rows_c <- sample(x = 1:nrow(crop_data), size = 0.70 * nrow(crop_data))
  train_c <- crop_data[rows_c, ]
  test_c  <- crop_data[-rows_c, ]
  
  # POOLED MODEL PERFORMANCE 
  # Use the big RF_pooled to predict on this crop's specific test set
  pred_pooled <- predict(RF_pooled, newdata = test_c)
  
  # CROP-SPECIFIC MODEL
  # Predictors excluding crop dummies (since they are constant for one crop)
  predictors_c <- c("Month", "Year", "CPI", "ttport_1", "ttcity_u5", "popdens", 
                    "bio_3", "bio_6", "bio_9", "bio_12", "bio_18","rain.sum.lag")
  
  # Tune specifically for this crop to be fair
  trf_c <- tuneRF(x = train_c[, predictors_c], y = train_c$pkg_real, stepFactor = 1.5, improve = 0.01, ntreeTry = 500, plot = FALSE, trace = FALSE) 
  
  best_mtry_c <- trf_c[which.min(trf_c[, 2]), 1]
  rf_specific <- randomForest(pkg_real ~ ., data = train_c[, c("pkg_real", predictors_c)],mtry = best_mtry_c, ntree = 1000, nodesize = 5)
  
  pred_specific <- predict(rf_specific, newdata = test_c)
  
  # METRICS 
  metrics <- data.frame(
    Crop = target_crop,
    RMSE_Pooled = round(sqrt(mean((test_c$pkg_real - pred_pooled)^2, na.rm=TRUE)), 2),
    R2_Pooled   = round(summary(lm(test_c$pkg_real ~ pred_pooled))$r.squared, 2),
    RMSE_Specific = round(sqrt(mean((test_c$pkg_real - pred_specific)^2, na.rm=TRUE)), 2),
    R2_Specific   = round(summary(lm(test_c$pkg_real ~ pred_specific))$r.squared, 2)
  )
  
  return(metrics)
}

```

```{r}
# Run and save results
all_crops <- unique(mypts$Crop)
comparison_results_tt <- do.call(rbind, lapply(all_crops, evaluate_models))

print(comparison_results_tt)
```

```{r}
comparison_df_wide3 <- comparison_results_tt %>%
  rename(
    RMSE_Pooled = RMSE_Pooled,
    R_squared_Pooled = R2_Pooled,
    RMSE_Crop_Specific = RMSE_Specific,
    R_squared_Crop_Specific = R2_Specific)
```

```{r}
knitr::kable(comparison_df_wide3, digits = 2, caption = "Table 3: Model Comparison (70/30 Train-Test Split Validation)")
```

```{r}
write.csv(comparison_df_wide3, "Train-Test-Validation-results.csv", row.names = FALSE)
```
 
# Comparison of regional and predicted rural maize prices for January 2025
```{r}
# Comparison of regional and predicted rural maize prices
# We use the deflated prices (pkg_real) for 2025
mypts_df <- as.data.frame(mypts) %>% 
  filter(Year == 2025, Crop == "Maize") %>%
  dplyr::select(Region_GADM, Month, pkg_real) 
```

```{r}
# The comparison loop
comparison_list <- list()

for (m in 1:12) {
  message("Processing month: ", m)
  
  # Load Prediction Raster (trained on deflated prices)
  pred_file <- sprintf("C:/Users/LMADAGA/OneDrive - CGIAR/H/Tanzania Price data/Datasets/Pred-plots/Maize/maize_price_rf_pred_%02d.tif", m)
  if (!file.exists(pred_file)) next
  
  pred_maize <- rast(pred_file) |> project(crs(tza1))
  names(pred_maize) <- "Predicted"
  
  # Calculate Regional Benchmark using pkg_real
  dta_reg <- mypts_df %>%
    filter(Month == m) %>%
    group_by(Region_GADM) %>%
    summarise(Reg_max_pkg = max(pkg_real, na.rm = TRUE)) %>%
    ungroup()
  
  # Apply Njombe patch
  if("Iringa" %in% dta_reg$Region_GADM & !"Njombe" %in% dta_reg$Region_GADM){
    iringa_val <- dta_reg$Reg_max_pkg[dta_reg$Region_GADM == "Iringa"]
    dta_reg <- bind_rows(dta_reg, tibble(Region_GADM = "Njombe", Reg_max_pkg = iringa_val))
  }
  
  # Spatial Join to Raster
  tza1_with_reg <- merge(tza1, dta_reg, by.x = "NAME_1", by.y = "Region_GADM", all.x = TRUE)
  regional_raster <- rasterize(tza1_with_reg, pred_maize, field = "Reg_max_pkg")
  
  # Data Extraction
  # Convert raster to dataframe and extract regional values
  tmp_df <- as.data.frame(pred_maize, xy = TRUE, na.rm = TRUE)
  
  # Get regional max values for each pixel coordinate
  reg_vals <- terra::extract(regional_raster, tmp_df[,c("x","y")])
  tmp_df$Regional_Max <- reg_vals[,2] # Extract the second column (the values)
  
  # Calculate Difference
  tmp_df <- tmp_df %>%
    mutate(
      Diff = Regional_Max - Predicted, 
      Month = m
    )
  
  # Attach Region Names for the X-axis
  pts_tmp <- vect(tmp_df, geom = c("x", "y"), crs = crs(tza1))
  tmp_df$Region <- terra::extract(tza1["NAME_1"], pts_tmp)[,2]
  
  comparison_list[[m]] <- tmp_df
}

comparison_all <- bind_rows(comparison_list) %>%
  filter(!is.na(Region), !is.na(Diff))

comparison_all$Month_Label <- factor(comparison_all$Month, levels = 1:12, labels = month.name)


```

```{r}
# Filter specifically for January
january_comparison <- comparison_all %>%
  filter(Month == 1) 

mrp <- ggplot(january_comparison, aes(x = Region, y = Diff)) +
  geom_boxplot(fill = "#A6CEE3", color = "black", outlier.size = 0.3, alpha = 0.7, size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.6) +
  # Note: facet_wrap is removed to focus on the single month
  labs(
    x = "",
    y = "Real Regional Price - Real Predicted Price (TZS/kg)",
    title = "",
    subtitle = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.title.y = element_text(size = 9),
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 9), # Tilted for better readability
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5))

mrp
```


## Prediction Evaluation
### Using 2025/2026 Validation data
```{r}
pred<-predict(object=rf, newdata=validation_data)
actual<-validation_data$pkg_real
result<-data.frame(actual=actual, predicted=pred)
```

```{r}
mse <- mean((actual - pred)^2, na.rm=TRUE)
paste('Mean Squared Error:', mse)
```

```{r}
rmse <- sqrt(mse)
paste('Root Mean Squared error: ',mean(sqrt(rf$mse)))
```

```{r}
#Save predicted & observed price
write.csv(result, "result.csv")
```

```{r}
#reading result.csv file (predicted vs observed)
rslt <- read.csv("result.csv", header=T)
print(names(rslt))
```

```{r}
#RMSE predicting from rf - predicited vs observed 
rf.rmse<-round(sqrt(mean( (rslt$actual-rslt$predicted)^2 , na.rm = TRUE )),2)
print(rf.rmse)
```

```{r}
#R-square
rf.r2<-round(summary(lm(actual~predicted, rslt))$r.squared,2)
print(rf.r2)
```

```{r}
range(actual)
```

```{r}
range(pred)
```

```{r}
#plotting predicted Vs observed
ggplot(result, aes(x=actual, y=predicted), alpha=0.6) +
  geom_point(colour = "blue", size = 1.4, alpha=0.6) +
  ggtitle('Random Forest "Wholesale Grain Prices in Tanzania"') +
  scale_x_continuous("Observed Price (Tsh) Per Kg",
                     limits = c(0, 5000),
                     breaks = seq(0, 5000, 1000)) +
  scale_y_continuous("Predicted Price (Tsh) Per Kg",
                     limits = c(0, 5000),
                     breaks = seq(0, 5000, 1000)) +
  theme(axis.line = element_line(colour = "black"),
        axis.text.y = element_text(size = 8, angle = 90, hjust = 0.5, vjust = 1),
        axis.text.x = element_text(size = 8)) +
  geom_abline(intercept = 0, slope = 1, linewidth = 0.5) +
  geom_smooth(aes(x = actual, y = predicted), formula = y ~ x, method = "lm", se = FALSE, colour = "red", linetype = 2, size = 0.9) +
  annotate("text", x = 300, y = 4500, label = paste("RMSE:", rf.rmse)) +
  annotate("text", x = 300, y = 4200, label = paste("R^2: ", rf.r2), parse = TRUE)

```


# Partial dependence plots
Plot Partial dependence of all the variables used except food commodities and months.
```{r}
library(caret)

var_importance <- varImp(rf)

impvar <- rownames(var_importance)[order(var_importance[, 1], decreasing=TRUE)]
```

```{r}
op <- par(mfrow=c(2, 4))
# exclude crop commodities and months
predictors_to_plot <- setdiff(impvar, c("maize", "rice", "sorghum", "bmillet", "fmillet", "wheat", "beans", "potato", "jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec"))

for (i in seq_along(predictors_to_plot)) {
  partialPlot(rf, as.data.frame(training_data), predictors_to_plot[i], xlab=predictors_to_plot[i],
              main="Partial Dependence")
}
```

Some descriptive statistics for each crop
```{r}
# Compute descriptive statistics for each crop
stats <- prices.monthly.long %>%
  filter(pkg_real > 0) %>%  
  group_by(Crop) %>%
  summarise(
    Mean = mean(pkg_real, na.rm = TRUE),
    Median = median(pkg_real, na.rm = TRUE),
    Minimum = min(pkg_real, na.rm = TRUE),
    Maximum = max(pkg_real, na.rm = TRUE),
    Std_Dev = sd(pkg_real, na.rm = TRUE),
    IQR = IQR(pkg_real, na.rm = TRUE),
    Observations = n()
  ) %>%
  arrange(Crop)

print(stats)
```


